Relationships in the extremes and their influence on competition and coexistence

### 2021-09-07 paper/Paper.Rnw
working on lit review in intro

%[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
%There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here  In one of the first empirical tests for storage effect and its requirements, \cite{Adler_2006} showed that climate variability - including extreme events - promoted the coexistence of grass species in Kansas prairie, specifically their ability to recover from low densities. Moreover, in studies on the invasibility of plants, the density and competitive balance of invasive Japanese knotweed increases as resource and thus climate variability increased.\citep{Parepa_2013} 
Environmental fluctuations are becoming more influenced by extreme climatic events \citep{Hansen_2012} as extreme events increase in frequency, duration, and intensity \citep{Meehl_2004, Ummenhofer_2017}. Extreme events (heat-waves, floods, hurricanes, etc.) alter the means and variation of environmental fluctuations, and thus have implications for coexistence and competition. Competitive output and community composition of grassland plants has been shown to be impacted by heat-waves and extreme drought \citep{White_2000, White_2001, Hoover_2014}. Studies on the direct effect of extreme events on competing species are slim (but see above); however, the consequences of change in environmental means, which may be due to extreme events or gradual climate change in general, on coexistence is more commonly studied.  \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Increased variability of environmental fluctuations is a newer consideration for studying environmental changes in comparison to changes in means (vitation); however, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

Extreme events can have important ecological impacts for controphic competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive (last longer, have higher duration) than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (similar, correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. This is a change in ATA not just from the event alone, but from the relationship between species. Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in gerneral (see above), they do not directly consider ATA between species response caused by environment (but see Ghosh for impact of ATA on community aggregate properties).

>>>>

%[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
%There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here  In one of the first empirical tests for storage effect and its requirements, \cite{Adler_2006} showed that climate variability - including extreme events - promoted the coexistence of grass species in Kansas prairie, specifically their ability to recover from low densities. Moreover, in studies on the invasibility of plants, the density and competitive balance of invasive Japanese knotweed increases as resource and thus climate variability increased.\citep{Parepa_2013}

Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}, influenced by extreme climatic events as they increase in frequency, intensity, duration, and exent \citep{Meehl_2004, Ummenhofer_2017}. 
The rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. are a product of climate change and have implications for coexistence and competition. Changes in means can be attributed to extreme events; however changes in means are moreso a result of gradual increases or decreases over time in an environmental variable. 
Extreme events push environmental variables to their extremes, increasing variance. Effects on competition and coexistence due to climate change are more widley researched through studing changes in mean environmental changes. 
For example, decreased precipitation decreased the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity as dominant species become excluded \citep{Mariotte_2013}. At the same time, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease.    
In contrast, increased environmental variance has been less commmonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Morever, in regard to invasive species, increase climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

Extreme events can have important ecological impacts for controphic competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. This is a change in ATA not just from the event alone, but from the relationship between species. Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in gerneral (see above), they do not directly consider ATA between species response caused by environment (but see Ghosh for impact of ATA on community aggregate properties).
